How to select predictive models for decision-making or causal inference?

### 2023-12-17 how_to_select_causal_models.tex
iter

Beyond rankings, the differences in terms of absolute
ability to select the best model are large: The R-risk selects a model
with a $\tau\text{-risk}$ only 1\% higher
than the best
possible candidate for strong overlap on Caussim, but selecting with
the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
--as in some causal-inference methods \citep{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
(Figure
\ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
datasets, the $R\text{-risk}$ consistently decreases the
risk compared to the $\mu\text{-risk}$:
0.1\% compared to 1\% on ACIC2016,  1\% compared to 20\% on ACIC2018,
and 0.05\% compared to 1\% on Twins.

@@

Model selection for causal inference becomes more and more difficult with
increasingly different treated and control populations (Figure
\ref{fig:all_datasets_overlap_effect_r_risk}). The absolute Kendall's
coefficient correlation with $\tau\text{-risk}$ drops from values around 0.9
(excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
(\ref{apd:experiments:additional_results}).

@@

Practically, we recommand a train/test split of 90 \%/10\%. Use most of the data
to estimate both the nuisances and candidates, then compute the risks on the
remaining test set for model selection (empirical argument in Appendix
\ref{apd:results:data_split}).

@@

For every risk, the oracle version recovers better the best estimator.
However,
stacked nuisances estimators (boosting and linear) lead to feasible
metrics with close performances to the oracles ones: the
corresponding estimators recover well-enough the true nuisances.
One may wonder if simpler models for the nuisance could be useful,
in particular in data-poor settings or when the true models are linear.
Figure \ref{fig:all_datasets_nuisances_comparison} compares causal model
selection estimating nuisances with stacked estimators or linear model.
It comprises the Twins data, where the true propensity model is linear,
and a downsampled version of this data, to study a situation favorable to
linear models. In these settings,
stacked and linear estimations of the nuisances performs equivalently.
Detailed analysis (Figure \ref{apd:fig:nuisances_comparison_twins})
confirms that using adaptive models --as built by
stacking linear models and gradient-boosted trees-- suffices to estimate nuisance.

>>>>

Beyond rankings, the differences in terms of absolute
ability to select the best model are large: The R-risk selects a model
with a $\tau\text{-risk}$ only 1\% higher
than the best
possible candidate for strong overlap on Caussim, but selecting with
the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
--as in some causal-inference methods \citep{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
(Figure
\ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
datasets, the $R\text{-risk}$ consistently decreases the
risk compared to the $\mu\text{-risk}$: from
0.1\% to 1\% on ACIC2016,  1\% from to 20\% on ACIC2018,
and 0.05\% from to 1\% on Twins.

@@

Model selection for causal inference becomes more and more difficult with
increasingly different treated and control populations (Figure
\ref{fig:all_datasets_overlap_effect_r_risk}). The absolute Kendall's
coefficient correlation with $\tau\text{-risk}$ drops from 0.9
(excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
(\ref{apd:experiments:additional_results}).

@@

Empirically, the best split is 90\,\%/10\,\%:
using 90\,\% of the data
to estimate both the nuisances and candidates, then computing the risks on the
remaining test set for model selection (experiments in Appendix
\ref{apd:results:data_split}).

@@

Stacked nuisances estimators (boosting and linear) lead to feasible
metrics with close performances to the oracles ones: the
corresponding estimators recover well-enough the true nuisances.
One may wonder if simpler models for the nuisance could be useful,
in particular in data-poor settings or when the true models are linear.
Figure \ref{fig:all_datasets_nuisances_comparison} compares causal model
selection estimating nuisances with stacked estimators or linear model.
It comprises the Twins data, where the true propensity model is linear,
and a downsampled version of this data, to study a situation favorable to
linear models. In these settings,
stacked and linear estimations of the nuisances performs equivalently.
Detailed analysis (Figure \ref{apd:fig:nuisances_comparison_twins})
confirms that using adaptive models --as built by
stacking linear models and gradient-boosted trees-- suffices to estimate nuisance.
